Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-03-24 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: reshaping content. To be continued.

World-leading research relies on software

``Today there are very few science areas left which do not rely on IT and thus
software for the majority of their research work. More importantly key
scientific advances in experimental and observational science would have been
impossible without better software.'' Kersten Kleese van Dam Pacific Northwestern
National Laboratory via change.org campaign

``Scientific discovery and innovation are advancing along fundamentally new
pathways opened by development of increasingly sophisticated software. Software
is an integral enabler of computation, experiment and theory, and directly
responsible for increased scientific productivity and enhancement of
researchers' capabilities.'' Dan Katz SI2 Program Director, National Science
Foundation.

We do not recognize these hundreds of thousands of researchers because we think
they are the long tail and forget that they are the mainstream and we are the
elite minority.

Software is not special; it is mainstream software

What would happen to your research without software

Survey of researchers from 15 Russell Group unis conducted by SSI between Aug- Oct 2014.
406 respondents covering representative range of funders, discipline and seniority.

This is not just about the ``traditional'' sciences

And 	attitudes 	are	changing, which 	is	changing 	the	situation

Survey of  PhD students from Russell Group university conducted by SSI between Aug -- Sep 2014.
approximately 100 respondents covering representative range of discipline.

Even	though 	a	lot	is	still	the	same

Survey of 138 University professionals conducted by SSI in Feb-Apr 2013.

The cost of UK research that relies on software

\textsterling 840 million of investment in 2013--2014 financial year, an amount that has risen by
3\% on average over last four years

30\% Of total research investment has been spent on research which relies on software
over the last four financial years

Analysis of data from 49,650 grant titles and abstracts published on Gateway to
Research covering 2010-2014.

The UK research community is just starting to understand the magnitude of the issue

56\% Of UK researchers develop their own research software

71\% Of UK researchers have had no formal software development training

140,000 UK researchers are relying on their own coding skills

4\% Of jobs advertised in UK universities were software related

Survey of researchers from 15 Russell Group unis conducted by SSI between Aug - Oct 2014.

If 92\% of researchers care about software and WSSSPE is about caring about
software---what are we going to do to help researchers?

Sustainability

With apologies to Greg Wilson:
\url{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}

We	can 	talk	all	day 	about 	the best 	ways	of	achieving sustainability
But	what	are	we	going 	to	do to	make	it	actually sustainable?

Careers outside academic sector: 
Non-university Research (industry, government, etc.)

>>>>

Neil Chue Hong is currently serving as the Director of the Software
Sustainability Institute (SSI), United Kingdom. Hong started with making the
point that world-leading research relies on software, quoting Kersten Kleese Van
Dam of the Pacific Northwestern National Laboratory via the campaign
\href{http://change.org}{change.org}, ``\emph{Today there are very few science
areas left which do not rely on IT and thus software for the majority of the
their research work. More importantly key scientific advances in experimental
and observational science would have been impossible without better software.}''
He also cited Daniel Katz, Sustained Innovation (SI2) Program Director of the
National Science Foundation, ``\emph{Scientific discovery and innovation are
advancing along fundamentally new pathways opened by development of increasingly
sophisticated software. Software is an integral enabler of computation,
experiment and theory, and directly responsible for increased scientific
productivity and enhancement of researchers' capabilities.}''

Hong drew attention to the misconception that we too often disregard hundreds of
thousands of software researchers thinking they are the long tail. We forget
that actually they are the mainstream and we are the elite minority. He
emphasized that software is no longer special; it is prevailing in scientific
research. Survey of researchers from 15 Russell Group universities conducted by
the SSI between Aug--Oct 2014 included a total of 406 respondents covering a
representative range of funders, disciplines, and seniority: 92\% confirmed the
use of research software such as MATLAB, SPSS, Excel, so on and so forth; 89\%
affirmed that it would be impossible or difficult to conduct research without
software. %Researchers often discover new software predominatly through
%recommendation of colleagues or web searches.
Nevertheless the British research community is just starting to understand the
magnitude of the issue. In the aforementioned SSI survey, 56\% of  
researchers develop their own research software; 71\%  have had
no formal software -development training; 140,000 researchers rely on
their own coding skills; 4\% of jobs advertised in British universities were software
related.

According to Hong, costs of software-reliant research in the United Kingdom include
\textsterling 840 million of investment in the financial year 2013--2014, an
amount that has risen by 3\% on average over the past four years. About 30\% of
total research investment has been spent on research which relies on software
over the last four financial years. These numbers stemmed from an analysis of
data from $49,650$ grant titles and abstracts published on Gateway to Research
covering years 2010--2014.

Hong led the audience to ponder on the questions that if 92\% of researchers and WSSSPE
care about scientific software, what are we going to \emph{do} to help researchers and 
to achieve software sustainability?
\begin{comment}
With apologies to Greg Wilson:
\href{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}
{You Keep Using That Word}
We can talk all day about the best ways of achieving sustainability
But what are we going to do to make it actually sustainable?
\end{comment}
Careers outside academic sector: 
Non-university Research (industry, government, etc.)
